---
title: "US Lilac R Code"
author: "Stephanie Nummer"
date: "November 27, 2018"
output: html_document
---

##Setting up front matter

  The front matter, aka loading the packages and setting the base directory, is important to ensure the ease of coding for the writer and other code readers. If someone wishes to run this code, the majority of changes in file directory will be made by changing it in the front matter. Additionally, in this chunk I am going to import the base dataset that will be used for all future coding and data managememt. 

```{r setup, include=FALSE}
packages<-function(x, repos="http://cran.r-project.org", ...){
  x<-as.character(match.call()[[2]])
  if (!require(x,character.only=TRUE)){
    install.packages(pkgs=x, repos=repos, ...)
    require(x,character.only=TRUE)
  }
}

#Setting base directory and data directory
base <- "C:/Users/steph/Documents/UToledo_PHD/PhD Research/HockeyStick Model"
dataDIR <- paste(base, "Data/Lilac", sep="/")
plotDIR <- paste(base, "manuscript", sep="/")
setwd(base)

#Loading Packages
packages(arm)
packages(grDevices)
packages(lattice)
packages(rv)
packages(rstan)
packages(tidyverse)
packages(dplyr)
packages(ggplot2)
packages(reshape2)
packages(tidyverse)
packages(cowplot)
packages(GGally)
packages(mapproj)
packages(maps)
packages(rgdal)
packages(rgeos)
packages(rstan)
packages(rstudioapi)
packages(maptools)
packages(RColorBrewer)
packages(ggmap)
packages(sp)
packages(gridExtra)

#importing data
USLilac <- read.csv(paste(dataDIR, "NAmlilac.csv", sep="/"), header=T)
```

##Cleaning Up Data & Formatting

  In the following chunck I will be performing data management to parse out the data needed to run the Bayesian hierarchcial hockey-stick model. The beginning of this cunck incorporated the clone type names associated with the phenotypes listed and changes the creators use of 999 as a missing value to NA. This is followed by importing the station specific information and performing the data management to match the site information with the phenological observation. This was done by first determining which stations had at least 30 years of data and then matching it with the latitude, longitude, evelvation, station name and state each station is located in. This matched data was then exported for later use in ArcGIS for matching based on geographic location. The final portion of this section is to add the station information to the dataset that has at least 30 years of information per station for use in the Bayesian hierarchcical hockey-stick model, summarizing the number of stations per phenotype in the resulting dataset, and testing a line of code that is later incorporated into the Stan model. 

```{r, message=FALSE}
#Adding Type column based on lilac species
USLilac$type <- "Syringa chinensis clone"
USLilac$type[USLilac$Ptype==2] <-"Syringa vulgaris"

#Changing 999 (their code for NA) to NAs
USLilac$FirstLeaf[USLilac$FirstLeaf==999] <- NA
USLilac$FirstBloom[USLilac$FirstBloom==999] <- NA

#Importing Station Data
USLilacST <- read.csv(paste(dataDIR, "NAmlilacsST.csv", sep="/"), header=T)

##Keeping data with at least 30 years worth of records
keep <- !is.na(match(USLilac$STID,
                     as.numeric(names(table(USLilac$STID))[table(USLilac$STID)>=30])))
uslilacs <- USLilac[keep,]
stlist <- unique(uslilacs$STID)
keep <- !is.na(match(USLilacST$ID, stlist))
sttype1 <- unique(USLilac$STID[USLilac$Ptype==1])
sttype2 <- unique(USLilac$STID[USLilac$Ptype==2])
uslilacsST <- USLilacST[keep,]

#Adding informtion about each station to the list of station we are keeping
US_Lilacs_StationInfo<-data.frame("Station ID"=uslilacsST$ID, "Latitude"=uslilacsST$Lat, "Longitude"=uslilacsST$X.Long, "Elevation"=uslilacsST$X.Elev, "Station Name"=uslilacsST$Station, "State"=uslilacsST$State)

#Writing information into a csv for later access and use in ArcGIS
write.csv(US_Lilacs_StationInfo, file="C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\Papers\\Publication Manuscripts\\Lilac Manuscript - Global Change Biology\\US Lilac Station Info.csv")

#Adding latitude and longitude information to dataframe for input to stan model
sort(uslilacsST$ID)
sort(unique(uslilacs$STID))
uslilacsST <- uslilacsST[order(uslilacsST$ID),]
uslilacs$Lat <- uslilacsST$Lat[as.numeric(ordered(uslilacs$STID))]
uslilacs$Lon <- uslilacsST$X.Long[as.numeric(ordered(uslilacs$STID))]
uslilacs$Elev <- uslilacsST$X.Elev[as.numeric(ordered(uslilacs$STID))]


## station lat/lon
sites <- data.frame(stid = levels(ordered(uslilacs$STID)),
                    lat = uslilacs$Lat[cumsum(table(uslilacs$STID))],
                    lon = uslilacs$Lon[cumsum(table(uslilacs$STID))],
                    elv = uslilacs$Elev[cumsum(table(uslilacs$STID))])


###Getting Ptype for each Station
StPtype<-uslilacs %>%
  group_by(STID) %>%
  summarise( Ptype=mean(Ptype))

#Testing code to stan model
GrYrMin<-tapply(uslilacs$Year, uslilacs$STID, min)

```

##US Lilac Map Data

  In this small section, I create a dataset meant to graph the stations by phenotype.

```{r, FALSE}

#Getting information arranged by phenotype to map later
USLilacMap<-USLilacST
USLilacMap$Type<-"PType 1"
temp <- !is.na(match(USLilacST$ID, sttype2))
USLilacMap$Type[temp]<-"PType 2"

```

##Bayesian Multilevel Model Lilac Model Using Step Function

  Below is the development of the Bayesian hierarchcial hockey-stick model in rstan. The hockey-stick model used below was first created as a function in R, to reduce the amount of time needed when performing this model on multiple datasets and to reduce the chance of user error in typing and copying. In this function the inputs needed are: the dataset, what the response variable is, the year, and the groupings (typically location, but may be another exchangable factor such as species). The first step of the function is to extract the different pieces of data needed and assign them a variable name. Next, the groupings are transitioned from a factor to a numeric variable and then the range of years for each grouping is determed. Essentially, the minumum and maxixmum year of phenological observation is calculated for each group and stored in a separate variable. The next step below is calculating the number of datapoints and the number of groups in the dataset. From there the data to put into the Stan code is compiled into a list. This list of input data includes the number of datapoints, the number of groups, the day of year of the first blooms (aka the response), the first year observed for each group, and the last year observed for each group. Next, the initial values are assigned for all parameters in the stan model. These values are assigned in a for loop to allow for a different random variable to start as the initial value of the model for each chain. Next, the lists creted thus far are compiled in to 1 list with the model code for input into rstan. 
  
  The model below follows the functional form of: 
  
$$y_{ij}=\{ \begin{array}
{rcl}
\alpha_j+\epsilon_{1j}, & & x_i < \phi_j \\
\alpha_j+\beta_j(x_i-\phi_j)+\epsilon_{2j}, & & x_i \ge \phi_j
\end{array}
$$
  
where where $y_i$ is the response measured in day of year, the subscript $\j$ represents location, $\alpha$ is the intercept of the left line segment (representing the average response before the effect of climate change on the response can be detected through the model), $\beta$ is the slope of the right line segment (representing the average annual change of the response since the effect of climate change on the response became detectable), $x_i$ is the calendar year, $\phi$ is the changepoint where the two line segments meet (the year when climate change effect on the response become detectable).The changepoint, $\phi$, for each location is limited by the first and last year of monitoring for that group. As such, the changepoint cannot be outside the range of data when monitoring occurred for each location. If the change occurred before the first year or after the last year of the record, the estimated changepoint would be concentrated on one or the other end of the time record. In addition, $\epsilon_1$ and $\epsilon_2$ are the error terms associated with the model (representing the annual variability of the response variable) and assumed to be normal random variables with a constant variance $\epsilon_1 \sim N(0, ??_{1,2})$ and $\epsilon_2 \sim N(0, ??_{2,2})$. This model specification allows the two error terms, $\epsilon_1$ and $\epsilon_2$, to have different variances. 

  To achieve this in rstan, I declared the data, parameters, transform 1 parameter, and model the response (day of year of first bloom) and the hyperparameters. Data in the Stan model include the integers with a lower threshold of 0 for the number of data points, the number of groups, and group ID's; the real value of the response (day of year) and predictor vairables year; and the vector with lengths equal to the number of groups of the first and last year of data collection. The parameters delaired in the model are the real hyperparameters of $\alpha$, $\phi$, and $\beta$, and the variance of the response, group-level $\alpha$, group-level $\phi$, and group-level $\beta$; and the vectors for the group-level $\alpha$, $\beta$, and raw $\phi$. 

  The parameter $\phi$ has varying bounds (the first and last year that observations were made) at the group level. To allow the bounds of $\phi$ to vary at the group level I had to bring in the bounds as 2 different data values -- the first year and the last year -- and estimate the parameter $\phi$ as value between 0 and 1. This value was then transformed in the transformed parameters portion of the Stan code using the first and last year data values to produce the group-level phi estimation that has varying boundaries. This works because the group level parameter was estimated between 0 and 1, aka phiG_raw, and then multiply that by the range of years for the specific group. That value is then added to the starting year (minimum value) to produce the $\phi$ for each individual group (aka station or location). The basis for this transformation was obtained form Stan User Guide. URL: https://mc-stan.org/docs/2_18/stan-users-guide/vectors-with-varying-bounds.html. 

  This gives us the changepoint or $\phi$ for each group or location, but we don't have the hyperparameter yet. The hyperparameter shows up when modeling the distirbution of the group level parameters. The group level distribution for $\phi$ was also estimated with a normal distribution with a mean being the hyperparameter $\phi$ and a variance with a positive value. Now, this presents an issue becuase the same parameter, phiG, was declared or modeled twice. This causes a non-fatal warning that for this to not cause an issue in the model, the log-Jacobian must be added to the liklihood function when modeling phiG with a normal distribution and determining the hyperparameter. The transformation of phiG_raw to phiG is a linear transformation. Becuase of this, the derivative of the transformation is a constant. The constant is -(maximumYear-minimumYear), which will not change the liklihood function of phiG. This means that although phiG is declaired twice, the transformation and likihood function, the hyperparameter still acts as a shrinkage estimator and is a meaningful estimation of the overall average. 
  
  In the model declaration of the Stan code I include the mean for each data point; the normal distributions of the group-level $\alpha$, $\beta$, and the transformed $\phi$; the model to calculate the mean for each data point; and the normal distirbution of the response with mean which was calculated in the model portion of the code and the variance which was delared in the parameters. The last portion of the model code not described is the calculation of the mean for each datapoint. This was calculated as the group-level $\alpha$ minus the group-level $\beta$, multiplied by the year minus the changepoint ($\phi$) if the year is greater than the changepoint ($\phi$). This if statement is included through the code int_step(). In this variation of the model the first portion is $\alpha$ - $\beta$, that negative $\beta$ indicates a downward slope and thus resulting postive values of $\beta$ means a negative slope, or an earlier response in the year, and negative values of $\beta$ mean a positive slope, or getting later in the year.
  
  The final portion of this chunk brings together the inputs needed for the MCMC sampling in stan, compiles the model code, samples and runs the model for 4 chains of 100,000 iterations, and saves the results as Rdata. 
  
```{r, message=FALSE}

packages(rstan)

hockey.step <- function(infile, variable, Year, Group, n.chains=4){
  y <- infile[,variable]
  temp <- !is.na(y)
  infile <- infile[temp,]
  y <- y[temp]
  x <- infile[,Year]
  

  Groupo<-infile[,Group]
  groupO <- ordered(Groupo)
  group <- as.numeric(groupO)
  
  ##Getting Min Year for Each Station
  GrYrMin<-tapply(infile[,Year], infile[,Group], min)
  GrYrMax<-tapply(infile[,Year], infile[,Group], max)
  

  n <- length(y)
  n.gr <- max(group)
  L <- GrYrMin
  U <- GrYrMax
  inits <- list()
  bugs.dat <- list(n=n, ngr=n.gr, y=y, x=x, group=group, minYr=L, maxYr=U)
  for (i in 1:n.chains)
    inits[[i]] <- list(alpha=rnorm(1), beta=runif(1), phi=runif(1), alphaG=rep(0, n.gr),
                       betaG=rep(0, n.gr), phiG=rep(0, n.gr),
                       sigmay=runif(1, 0, 3), sigmaA=runif(1), sigmaB=runif(1),sigmaP=runif(1))
  parameters <- c("alpha","beta", "phi", "alphaG", "betaG", "phiG", "sigmay", "sigmaA", "sigmaB", "sigmaP")
  return(list(para=parameters, data=bugs.dat, inits=inits,
              n.chains=n.chains, groupID=levels(groupO),
              model="
 /*  multilevel hockey stick model with group predictors */
data{
  int<lower=0> n;
  int<lower=0> ngr;
  real y[n];
  int<lower=0> group[n]; // grouping IDs
  real x[n]; // year
  vector[ngr] minYr; //lower bounds
  vector[ngr] maxYr; //upper bounds
}

parameters{
  real alpha; // intercept at change point
  real beta; // slope after change point
  real phi; // change point

  vector[ngr] alphaG;
  vector[ngr] betaG;
  vector<lower=0, upper=1>[ngr] phiG_raw;

  real<lower=0,upper=100> sigmay;

  real<lower=0> sigmaA;
  real<lower=0> sigmaB;
  real<lower=0> sigmaP;
}

transformed parameters {
  vector[ngr] phiG = minYr + (maxYr-minYr).*phiG_raw;
}

model{
      real mu[n];//mu for each observation

      alphaG ~ normal(alpha,sigmaA);
      betaG ~ normal(beta,sigmaB);
      phiG ~ normal(phi,sigmaP);

	for (i in 1:n){
    mu[i]=(alphaG[group[i]]) - (betaG[group[i]]) * int_step(x[i]-(phiG[group[i]]))*(x[i]-(phiG[group[i]]));
}
	    y~normal(mu, sigmay);

}
"
              ))
}


##editing chain information for model
rstan_options(auto_write = TRUE)
options(mc.cores = min(c(parallel::detectCores(), 8)))

nchains <-  min(c(parallel::detectCores(), 8))
niters <- 100000
nkeep <- 2500
nthin <- ceiling((niters/2)*nchains/nkeep)

input.to.stan <- hockey.step(infile=uslilacs, variable="FirstBloom", Year="Year", Group="STID", n.chains=4)
n.chains <- 4
n.iter <- 100000
n.keep <- 2000
nthin <- max(c(1,floor((n.iter)*n.chains/n.keep)))

comp_model<- stan_model(model_code=input.to.stan$model)

fit2 <- sampling(comp_model, data=input.to.stan$data,
             init=input.to.stan$inits,
             pars=input.to.stan$para,
             iter=niters, chains=input.to.stan$n.chains,
             thin=nthin)

packages(rstudioapi)

  print(fit2)
  save ("fit2", file=paste("11.15.19.AllStFit.LU.Rdata", sep=""))
  
```

##Calculating Coef. from Bayesian Results

  Below, the Rdata saved from the previous chunk is extracted and used to produce the mean values for each parameter and the 97.5% and 2.5% credible intervals of the hyperparameters and group-level coefficents for all groups/stations in the model. To produce the means and credible intervals of the hyperparameters and group-level coefficients from the Rdata I used the package rv and the function rvsims. This data is then combined with the site specific information to produce a dataset with the values for each hyperparameter with credible intervals, group-level coefficients with hyperparameters, site id, elevation, latitude, longitude, and phenotype for future geographic comparisons.

```{r, message=FALSE}

load("11.15.19.AllStFit.LU.Rdata")
fit2coefGr<- rstan::extract(fit2, permuted=T)

PhHypRv<-rvsims(fit2coefGr$phi)
AHypRv<-rvsims(fit2coefGr$alpha)
BHypRv<-rvsims(fit2coefGr$beta)

PhHyp<-as.data.frame(print(PhHypRv))
AHyp<-as.data.frame(print(AHypRv))
BHyp<-as.data.frame(print(BHypRv))

PhGrRv<-rvsims(fit2coefGr$phiG)
AGrRv<-rvsims(fit2coefGr$alphaG)
BGrRv<-rvsims(fit2coefGr$betaG)

PhGr<-as.data.frame(print(PhGrRv))
AGr<-as.data.frame(print(AGrRv))
BGr<-as.data.frame(print(BGrRv))

ODyear<-data.frame(Group=paste("Group", 1:53, sep=" "))

ODyear$stid<-sites$stid
##Customize to dataset
ODyear$lat<-sites$lat
##Customize to dataset
ODyear$lon<-sites$lon
##Customize to dataset
ODyear$elv<-sites$elv
##Customize to dataset
ODyear$Ptype<-StPtype$Ptype

ODyear$PhHyMean<-PhHyp$mean
ODyear$PhHy2.5<-PhHyp$`2.5%`
ODyear$PhHy97.5<-PhHyp$`97.5%`

ODyear$AHyMean<-AHyp$mean
ODyear$AHy2.5<-AHyp$`2.5%`
ODyear$AHy97.5<-AHyp$`97.5%`

ODyear$BHyMean<-BHyp$mean
ODyear$BHy2.5<-BHyp$`2.5%`
ODyear$BHy97.5<-BHyp$`97.5%`

ODyear$PhGrMean<-PhGr$mean
ODyear$PhGr2.5<-PhGr$`2.5%`
ODyear$PhGr97.5<-PhGr$`97.5%`

ODyear$AGrMean<-AGr$mean
ODyear$AGr2.5<-AGr$`2.5%`
ODyear$AGr97.5<-AGr$`97.5%`

ODyear$BGrMean<-BGr$mean
ODyear$BGr2.5<-BGr$`2.5%`
ODyear$BGr97.5<-BGr$`97.5%`

FByearST<-ODyear
##Customize to dataset

##Ordering things
as.vector(FByearST$lat)
##Change thigns to character strings with no vectors
pV<-as.vector(FByearST$lat)
pV<-ordered(pV)
##is now alphabetically ordered
pV.mu<-tapply(FByearST$PhGrMean, pV, mean)
##put things in order from least to greatest 
oo<-order(pV.mu)
oo
##in alphabetical
  levels (pV)
##puts in order of oo - so means going from least to greatest 
pV<-ordered(pV, levels=levels(pV)[oo])
pV
##from least to greatest

##FOR Beta
##Change thigns to character strings with no vectors
bV<-as.vector(FByearST$lat)
bV<-ordered(bV)
##is now alphabetically ordered
bV.mu<-tapply(FByearST$BGrMean, bV, mean)
##put things in order from least to greatest 
oo<-order(bV.mu)
oo
##in alphabetical
levels (bV)
##puts in order of oo - so means going from least to greatest 
bV<-ordered(bV, levels=levels(bV)[oo])
bV
##from least to greatest

FByearST$pV<-pV
FByearST$bV<-bV

FByearST$stid.lat<- reorder(FByearST$stid, FByearST$lat)

```

##Mapping all 53 Stations

  Below I map the resulting 53 stations with at least 30 years of data onto a map of North America using ggplot. I also include the 2 station worth noting that have the earliest and latest changepoints. 

```{r, message=FALSE}
#Getting information on 2 key stations noted in manuscript
stat2.a<-FByearST[FByearST$stid==49122,]
##A is blue 
stat2.b<-FByearST[FByearST$stid==213303,]
##B is purple

#Ensuring the latitude and longitude are in numeric format for mapping
stat2a.lat<-as.numeric(stat2.a$lat)
stat2a.long<-as.numeric(stat2.a$lon)

stat2b.lat<-as.numeric(stat2.b$lat)
stat2b.long<-as.numeric(stat2.b$lon)

#Mapping all stations with at least 30 years of data onto a map of the US and Canada and adding the 2 noted stations
ggplot(FByearST, aes(lon, lat))+borders(database="world", regions ="USA", xlim=c(-126, -50), ylim=c(28,50))+geom_point(shape=1, size=1.5)+ geom_point(aes(stat2a.long,stat2a.lat), shape=5, size=2, color="blue")+ geom_point(aes(stat2b.long,stat2b.lat), shape=2, size=2, color="purple")+theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), axis.line = element_blank())
ggsave(file="LilacSpeciesMap_53StationsSM.eps", device = "eps", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 3.3, height = 2.2)
dev.off()

```

##Plotting Results for Stations

  In this section I plot the hockey-stick model for each individual station and the overall average model (aka the model represented by the hyperparameters). This model is plotted using the general plot functions in r. I then use ggplot to take a deeper look at the $\phi$ and $\beta$ for each individual station. In these plots they are organized from the highest latitude at the top to the lowest latitude at the bottom of the graph. It is important to note that I cannot directly compare the credible intervals of individual stations to discern a significant difference because of the correlations among the estimated site-specific changepoints (Qian 2012), but I can compare the mean values for each station. 

```{R, message=FALSE}
#Plotting the hockey-stick graphs based on the determined parameters for all stations and the hyperparameters

#jpeg(file="C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU\\LilacMultilevelwHype.jpg")
postscript(file="C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU\\LilacMultilevelwHyperSM.eps",horiz=FALSE,onefile=FALSE,width=3.3,height=3.025,paper="letter")
plot(FirstBloom~Year,type="n", data=uslilacs, col="gray92", ylab="First Bloom Day of Year")
for (i in 1:53)
  segments(x0=c(1956, FByearST[i,16]), y0=c(FByearST[i,19], FByearST[i,19]), x1=c(FByearST[i,16],2003), y1=c(FByearST[i,19], FByearST[i,19]-(FByearST[i,22]*(2003-FByearST[i,16]))), col="gray80")
segments(x0=c(1956, FByearST[1,7]), y0=c(FByearST[1,10], FByearST[1,10]), x1=c(FByearST[1,7],2003), y1=c(FByearST[1,10], FByearST[1,10]-(FByearST[1,13]*(2003-FByearST[1,7]))), col="black", lwd=2)
dev.off()

#plotting the changepoint and associated credible intervals for all 53 stations based on latitude 
x<-print(ggplot(data=FByearST, aes(x = stid.lat, y = PhGrMean, ymax = PhGr97.5, ymin = PhGr2.5)) + geom_point(position=position_dodge(width=0.2), color="darkblue") + geom_errorbar(position=position_dodge(width = 0.2), width=0.1, color="darkblue")+coord_flip()+labs(x=("Station"), y=(expression(phi))))
ggsave(file="LilacMLevelPhiSTwHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 4, height = 10)
dev.off()

#plotting the slope after the changepoint and associated credible intervals for all 53 stations based on latitude
y<-print(ggplot(data=FByearST, aes(x = stid.lat, y = BGrMean, ymax = BGr97.5, ymin = BGr2.5)) + geom_point(position=position_dodge(width=0.2), color="darkblue") + geom_errorbar(position=position_dodge(width = 0.2), width=0.1, color="darkblue")+coord_flip()+labs(x=("Station"), y=(expression(beta))))
ggsave(file="LilacMLevelBetaSTwHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 4, height = 10)
dev.off()

#combining the slope after changepoint and changepoint graphs into 1 image
togST<-ggdraw() +
    draw_plot(x, x = 0, y = 0.02, width = 0.5, height =0.94)+
    draw_plot(y, x = 0.5, y = 0.02, width = 0.5, height =0.94) 
togST
ggsave(file="LilacMLM_STwHypeSM.eps", device = "eps", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 6.85, height = 8.5625)

```

##Plotting Station Phi and Beta Versus Latitude 

  Here I begin to explore the $\beta$ and $\phi$ for each station to see if there is a geographic correlation to to the values. In this chunk I plot the Values by latitude. 

```{r, message=FALSE}
#Turning tha latitudes into a vector so they may be organized
Stat.Lat.Rv<-as.vector(FByearST$lat)

Phi.Gr.Lat.Rv<-cc(PhGrRv, Stat.Lat.Rv)

#Plotting all 53 stations with by latitude with the latitude as the Y axis title for phi
LatPhi<-print(ggplot(data=FByearST, aes(x =lat, y = PhGrMean, ymax = PhGr97.5, ymin = PhGr2.5)) + geom_point(color="darkblue") + geom_errorbar(width=0.1, color="darkblue")+labs(x=("Latitude"), y=(expression(phi)))) +coord_flip()
ggsave(file="LilacMLevelPhiLatitudewHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 4, height = 8)
dev.off()

#Plotting all 53 stations with by latitude with the latitude as the Y axis title for beta
LatBeta<-print(ggplot(data=FByearST, aes(x =lat, y = BGrMean, ymax = BGr97.5, ymin = BGr2.5)) + geom_point(color="darkblue") + geom_errorbar(width=0.1, color="darkblue")+labs(x=("Latitude"), y=(expression(beta))))+coord_flip()
ggsave(file="LilacMLevelBetaLatitudewHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 4, height = 8)
dev.off()

#combining the 2 previous graphs
togLat<-ggdraw() +
    draw_plot(LatPhi, x = 0, y = 0.02, width = 0.5, height =0.94)+
    draw_plot(LatBeta, x = 0.5, y = 0.02, width = 0.5, height =0.94) 
togLat
ggsave(file="LilacMLM_LatwHype.eps", device = "eps", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 8, height = 8)

```

##Plotting Station Phi and Beta Versus Elevation

  In this chunk I continue the exploration of geographic trends to $\phi$ and $\beta$ by plotting these values against the station's elevations. 
  
```{r, message=FALSE}
#Plotting phi by elevation 
ElePhi<-print(ggplot(data=FByearST, aes(x =elv, y = PhGrMean, ymax = PhGr97.5, ymin = PhGr2.5)) + geom_point(color="darkblue") + geom_errorbar(width=0.1, color="darkblue")+labs(x=("Elevation"), y=(expression(phi))))+coord_flip()
ggsave(file="LilacMLevelPhiElevationwHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 4, height = 8)
dev.off()

#Plotting beta by elevation
EleBeta<-print(ggplot(data=FByearST, aes(x =elv, y = BGrMean, ymax = BGr97.5, ymin = BGr2.5)) + geom_point( color="darkblue") + geom_errorbar(width=0.1, color="darkblue")+labs(x=("Elevation"), y=(expression(beta))))+coord_flip()
ggsave(file="LilacMLevelBetaElevationwHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 4, height = 8)
dev.off()

#combining the 2 previous graphs
togEle<-ggdraw() +
    draw_plot(ElePhi, x = 0, y = 0.02, width = 0.5, height =0.94)+
    draw_plot(EleBeta, x = 0.5, y = 0.02, width = 0.5, height =0.94) 
togEle
ggsave(file="LilacMLM_ElewHype.eps", device = "eps", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 8, height = 8)

```

##Plotting Station Phi and Beta Versus PType

Below I investigate if there is a correlation trend between the $\phi$ and $\beta$ and phenotype of the lilac located at the station by plotting these values against the station's phenotype. 

```{r, message=FALSE}
#ensuring phenotype is a factor of 1 or 2 depending on clone type of lilacs
FByearST$Ptype<-as.numeric(FByearST$Ptype)

#Plotting phi by phenotype/clone type
PTypePhi<-print(ggplot(data=FByearST, aes(x =jitter(Ptype), y = PhGrMean, ymax = PhGr97.5, ymin = PhGr2.5)) + geom_point(color="darkblue") + geom_errorbar(width=0.05, color="darkblue")+labs(x=("Phenotype"), y=(expression(phi)))+scale_x_continuous(breaks = c(1,2))+coord_flip())
ggsave(file="LilacMLevelPhiPtypewHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 4, height = 8)
dev.off()

#Plotting beta by phenotype/clone type
PTypeBeta<-print(ggplot(data=FByearST, aes(x =jitter(Ptype), y = BGrMean, ymax = BGr97.5, ymin = BGr2.5))+ labs(x=("Phenotype"), y=(expression(beta))) + geom_point( color="darkblue") + geom_errorbar(width=0.05, color="darkblue")+scale_x_continuous(breaks = c(1,2))+coord_flip())
ggsave(file="LilacMLevelBetaPtypewHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 4, height = 8)
dev.off()

#combining both graphs into 1 
togPType<-ggdraw() +
    draw_plot(PTypePhi, x = 0, y = 0.02, width = 0.5, height =0.94)+
    draw_plot(PTypeBeta, x = 0.5, y = 0.02, width = 0.5, height =0.94) 
togPType
ggsave(file="LilacMLM_PTypewHype.eps", device = "eps", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 8, height = 6)

```

##Plotting Station Phi and Beta by Ecoregion

####Ecoregions were derived in GIS using an Ecoregion Level 1 Shapefile and Latitude and Longitude Data

Here I plot the $\beta$ and $\phi$ against the level 1 ecoregions that each station is located in. The ecoregions were matched to the latitude and longitude with ArcGIS. 

```{r, message=FALSE}

##save the FByearST dataframe as csv to use in gis
LilacFirstBloom_MLM_53_Results<-as.data.frame(FByearST)
write.csv(LilacFirstBloom_MLM_53_Results, paste(dataDIR, "LilacFirstBloom_MLM_Results.csv", sep="/"))

##map of the 53 stations
ggplot(FByearST, aes(lon, lat, color=as.character(Ptype)))+borders(database="world", regions = c("USA","Canada"), xlim=c(-126, -50), ylim=c(28,50))+geom_point(shape=1)+theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), axis.line = element_blank()) + scale_colour_discrete(name = "Phenotype")
ggsave(file="LilacSpeciesMap_53Stations.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 6, height = 4)
dev.off()

#importing the ecoregions by station csv
EcoregionStation<-read.csv(paste(dataDIR, "EcoregionsStation.csv", sep="/"), header=T)

#matching ecoregion information with results from model
FByearST$EcoregionNum<-EcoregionStation$NA_L1CODE

FByearST$EcoregionName<-EcoregionStation$NA_L1NAME

FByearST$EcoregionNum.jit<-jitter(FByearST$EcoregionNum)

#plotting phi for each station by ecoregion
EcoRegPhi<-print(ggplot(data=FByearST, aes(x = EcoregionNum.jit, y = PhGrMean, ymax = PhGr97.5, ymin = PhGr2.5)) + geom_point(aes(color=EcoregionName)) + geom_errorbar(aes(color=EcoregionName), width=0.05)+labs(x=("Ecoregion"), y=(expression(phi)))+scale_x_continuous(breaks = c(1,2))+coord_flip())
ggsave(file="LilacMLevelPhiEcoRegwHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 8, height = 8)
dev.off()

#editing legend for combination
EcoRegPhi2<-EcoRegPhi+theme(legend.position = "none")

#plotting phi for each station based on ecoregion as organized by latitude
EcoRegLatPhi<-print(ggplot(data=FByearST, aes(x =lat, y = PhGrMean, ymax = PhGr97.5, ymin = PhGr2.5)) + geom_point(color="darkblue") + geom_errorbar(width=0.1, color="darkblue")+labs(x="Latitude", y=(expression(phi))) + facet_grid(~EcoregionName) +coord_flip())
ggsave(file="LilacMLevelPhiEcoRegLatwHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 24, height = 8)
dev.off()

#plotting beta for each station by ecoregion
EcoRegBeta<-print(ggplot(data=FByearST, aes(x=EcoregionNum.jit, y = BGrMean, ymax = BGr97.5, ymin = BGr2.5, color=EcoregionName))+ labs(x=("Ecoregion"), y=(expression(beta))) +scale_x_continuous(breaks = c(1,2))+coord_flip()+geom_point()+ geom_errorbar(width=0.05))
ggsave(file="LilacMLevelBetaEcoRegwHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 8, height = 8)
dev.off()

#editing legend for combination
EcoRegBeta2<-EcoRegBeta + theme(legend.title = element_text(size = 10),
  legend.text = element_text(size = 6), legend.key.size = unit(0.5, "cm"), legend.key.width = unit(0.5,"cm"))

#plotting beta for each station based on ecoregion as organized by latitude
EcoRegLatBeta<-print(ggplot(data=FByearST, aes(x =lat, y = BGrMean, ymax = BGr97.5, ymin = BGr2.5)) + geom_point(color="darkblue") + geom_errorbar(width=0.1, color="darkblue")+labs(x="Latitude", y=(expression(beta))) + facet_grid(~EcoregionName) +coord_flip())
ggsave(file="LilacMLevelBetaEcoRegLatwHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 24, height = 8)
dev.off()

#plotting the graphs of phi and beta based on ecoregion together
togEcoReg<-ggdraw() +
    draw_plot(EcoRegPhi2, x = 0, y = 0.02, width = 0.33, height =0.94)+
    draw_plot(EcoRegBeta2, x = 0.33, y = 0.02, width = 0.67, height =0.94) 
togEcoReg
ggsave(file="LilacMLM_EcoRegwHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 12, height = 8)

#plotting the graphs of phi and beta based on ecoregion and latitude together
togEcoRegLat<-ggdraw() +
    draw_plot(EcoRegLatPhi, x = 0, y = 0, width = 1, height =0.5)+
    draw_plot(EcoRegLatBeta, x = 0, y = 0.5, width =1, height =0.5) 
togEcoRegLat
ggsave(file="LilacMLM_EcoRegLatwHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 24, height = 16)

```

##Plotting Station Phi and Beta by USDA Harshness Zone

####USDA Hardiness Zones were derived in GIS using an USDA Harshness Zone Shapefile and Latitude and Longitude Data

Lastly, the $\phi$ and $\beta$ are compared to the USDA Hardiness Zone based on latitude and longitude. 

```{r, message=FALSE}
#importation USDA Hardiness Zone Information as retrieved by GIS
USDAStation<-read.csv(paste(dataDIR, "StationUSDAHarshLocations.csv", sep="/"), header=T)

#Adding USDA Hardiness information to the model results
FByearST$USDAZone<-USDAStation$ZONE
FByearST$USDACode<-USDAStation$GRIDCODE
FByearST$USDACode.jit<-jitter(FByearST$USDACode)

#Graphing the phi for each station by hardiness zone
ZonePhi<-print(ggplot(data=FByearST, aes(x =USDACode.jit, y = PhGrMean, ymax = PhGr97.5, ymin = PhGr2.5)) + geom_point(aes(color=USDAZone)) + geom_errorbar(aes(color=USDAZone), width=0.05)+labs(x=("USDA Harshness Zone"), y=(expression(phi)))+scale_x_continuous(breaks = c(1,2))+coord_flip())
ggsave(file="LilacMLevelPhiUSDAZonewHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 8, height = 8)
dev.off()

#editing legend for combination
ZonePhi2<-ZonePhi+theme(legend.position = "none")

#Graphing the beta for each station by hardiness zone
ZoneBeta<-print(ggplot(data=FByearST, aes(x =USDACode.jit, y = BGrMean, ymax = BGr97.5, ymin = BGr2.5))+ labs(x=("USDA Harshness Zone"), y=(expression(beta))) + geom_point(aes(color=USDAZone)) + geom_errorbar(aes(color=USDAZone), width=0.05)+scale_x_continuous(breaks = c(1,2))+coord_flip())
ggsave(file="LilacMLevelBetaUSDAZonewHype.jpg", device = "jpg", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 8, height = 8)
dev.off()

#editing legend for combination
ZoneBeta2<-ZoneBeta + theme(legend.title = element_text(size = 10),
  legend.text = element_text(size = 6), legend.key.size = unit(0.65, "cm"), legend.key.width = unit(0.75,"cm")) 

#combining the beta and phi graphs as prganized by hardiness zones
togZone<-ggdraw() +
    draw_plot(ZonePhi2, x = 0, y = 0.02, width = 0.43, height =0.94)+
    draw_plot(ZoneBeta2, x = 0.43, y = 0.02, width = 0.57, height =0.94) 
togZone
ggsave(file="LilacMLM_ZonewHype.eps", device = "eps", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 12, height = 8)

```

###Plot Zones, Ecoregions, Latitude, Elevation Together

The section below brings the 4 geographic characteristics associated with station location graphed previously into 1 graph for the online supplemental material. 

```{r, message=False}
#Graphing all 4 graphs of phi and beta for each different characteristic together into 1 graph
togAll4<-ggdraw() +
    draw_plot(togLat, x = 0, y = 0.55, width = 0.475, height =0.423)+
    draw_plot(togEle, x = 0.52, y = 0.55, width = 0.475, height =0.423) +
    draw_plot(togEcoReg, x = 0, y = 0.02, width = 0.475, height =0.523)+
    draw_plot(togZone, x = 0.52, y = 0.02, width = 0.475, height =0.523) 

togAll4

#Saving graph
ggsave(file="LilacMLM_All4Geo.eps", device = "eps", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures\\Lilac.Manuscript.Updated.Beta.Phi.LU",  width = 16, height = 12)


```

##General Hockey-Stick Model Graph

The code below first creates a generic dataset with 3 value to use as endpoints to create a basic graph depicting the shape of the hockey-stick model. Then the dataframe was input into a ggplot graph with the dataframe used as endpoints for the 2 line segments making up a hockey-stick model. 

```{r, message=FALSE}
#Creating a temporary data frame to show a general hockeystick model graph
genDF<-data.frame(Y.int=c(1,1,0.005), x=c(0.005, 1, 2))

#Graphing the general hockeystick model graph with labels
genModel<-print(ggplot(data=genDF, aes(x, Y.int)) + geom_line(size=1) + scale_x_continuous(name="Year", labels=NULL, limits =c(0,2)) + scale_y_continuous(name="Response", labels=NULL, limits =c(0,1.5))+theme(axis.ticks = element_blank()) + annotate( "text", x=0.002, y=1.1, label=expression(alpha), color="blue", size=4)+ annotate( "text", x=1.04, y=1.07, label=expression(phi), color="dark green", size=4) + annotate( "text", x=1.65, y=0.5, label=expression(beta), color="purple", size=4))
ggsave(file="GeneralHockeyStickModelGraph.eps", device = "eps", path = "C:\\Users\\steph\\Documents\\UToledo_PHD\\PhD Research\\HockeyStick Model\\RMds\\Figures",  width = 3.3, height = 2.0625)

```


